\large Differentiable Cosmological Simulation with the Adjoint Method \vspace{0.3em}

### 2024-02-07 docs/papers/adjoint/adjoint.tex
Show hidden appendices

\usepackage{CJK}
\usepackage{microtype}
\usepackage{savesym}
  \savesymbol{lambdabar}
\usepackage{newpxtext, eulerpx}
  \restoresymbol{newpx}{lambdabar}
\usepackage[T1]{fontenc}
\usepackage{fontawesome}
\usepackage{amsmath}
%  \allowdisplaybreaks
\usepackage{bm}
\usepackage{nicefrac}
\usepackage[caption=false]{subfig}
\usepackage[figure,figure*]{hypcap}
\usepackage{tikz}
  \usetikzlibrary{positioning, fit, calc, arrows.meta}
\usepackage{booktabs}

@@

%\vspace{1em}
%\section{Hamiltonian Dynamics}
%
%Consider a Hamiltonian system, with $\vz = (\vx, \vp)$ where $\vx$ and
%$\vp$ are the canonical coordinates and momenta.
%Other variables split likewise: $\vDelta = (\vDelta_\vx, \vDelta_\vp)$,
%$\vLambda = (\vLambda_\vx, \vLambda_\vp)$, and $\vlambda =
%(\vlambda_\vx, \vlambda_\vp)$.
%Hamilton's equation of motion is
%%
%\begin{equation}
%\vf = \begin{pmatrix}
%  \hphantom{-} \p\cH/\p \vp \\ - \p\cH/\p \vx
%\end{pmatrix}.
%\label{Hamilton}
%\end{equation}
%%
%The dynamics is time-reversible: flipping the arrow of time $\d/\d t \to
%- \d/\d t$ and the direction of momenta $\vp \to -\vp$ preserve the form
%of \eqref{ode}.
%It can be reversible even after discretization if appropriate integrator
%is used.
%
%The variational equation \eqref{variational} and the adjoint equation
%\eqref{adjoint} become
%%
%\begin{align}
%\begin{pmatrix}
%  \dot\vDelta_{\vx\vx} & \dot\vDelta_{\vx\vp} \\
%  \dot\vDelta_{\vp\vx} & \dot\vDelta_{\vp\vp}
%\end{pmatrix}
%&= \begin{pmatrix}
%  \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vx}
%  & \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vp} \\
%  - \dfrac{\p^2\cH}{\p\vx\p\vx}
%  & - \dfrac{\p^2\cH}{\p\vx\p\vp} \\
%\end{pmatrix}
%\cdot
%\begin{pmatrix}
%  \vDelta_{\vx\vx} & \vDelta_{\vx\vp} \\
%  \vDelta_{\vp\vx} & \vDelta_{\vp\vp}
%\end{pmatrix}, \nonumber \\
%%
%\qquad
%- \begin{pmatrix}
%  \dot\vLambda_{\vx\vx} & \dot\vLambda_{\vx\vp} \\
%  \dot\vLambda_{\vp\vx} & \dot\vLambda_{\vp\vp}
%\end{pmatrix}
%&= \begin{pmatrix}
%  \hphantom{-} \dfrac{\p^2\cH}{\p\vx\p\vp}
%  & - \dfrac{\p^2\cH}{\p\vx\p\vx} \\
%  \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vp}
%  & - \dfrac{\p^2\cH}{\p\vp\p\vx} \\
%\end{pmatrix}
%\cdot
%\begin{pmatrix}
%  \vLambda_{\vx\vx} & \vLambda_{\vx\vp} \\
%  \vLambda_{\vp\vx} & \vLambda_{\vp\vp}
%\end{pmatrix}.
%\end{align}
%%
%Interestingly, their forms are identical once we swap rows of
%$\vLambda$, and flip the directions of time and momenta.
%
%As discussed before, $\vlambda$ follows adjoint equations similar to
%that of $\vLambda$ above.
%One can integrate it and \eqref{Hamilton} backward together to compute
%$\d\cJ / \d\vtheta$.
%
%
%\vspace{1em}
%\section{\texorpdfstring{$N$}{N}-body Dynamics}
%
%We model self-gravitating collisionless dark matter with $N$ particles
%each of mass $m$ in a periodic box of comoving volume $V$, whose
%physical length expands proportional to a scale factor $a(t)$.
%The canonical coordinates are the comoving positions $\vx = (\vx_1,
%\cdots, \vx_N)$, and their canonical momenta are $\vp = (\vp_1, \cdots,
%\vp_N)$ with $\vp_i = a^2 m \dot\vx_i$ for each particle.
%
%The $N$-body Hamiltonian is
%%
%\begin{equation}
%\cH = \sum_i \frac{\vp_i^2}{2 m a^2}
%+ \sum_{i<j} \frac{m^2 \varphi(\vx_i - \vx_j)}{a},
%\end{equation}
%%
%where $\varphi$ is the potential of a unit-mass particle at $a=1$ under
%periodic boundary condition
%%
%\begin{equation}
%\nabla^2 \varphi
%= 4\pi G \bigl[ \deltaD(\vx - \vx') - V^{-1} \bigr].
%\end{equation}
%%
%So the only nonzero second $\cH$ derivatives are
%%
%\begin{equation}
%\frac{\p^2\cH}{\p\vp_i\p\vp_j} = \frac{\vI_3}{2 m a^2},
%\qquad
%\frac{\p^2\cH}{\p\vx_i\p\vx_j}
%= \deltaK_{ij} \sum_{k \neq i} \frac{m^2}a
%  \frac{\p\varphi(\vx_i-\vx_k)}{\p\vx_i\p\vx_i}
%+ \frac{m^2}a \frac{\p\varphi(\vx_i-\vx_j)}{\p\vx_i\p\vx_j},
%\end{equation}
%%
%where $\vI_3$ is the $3\times3$ identity matrix, the sum is the tidal
%tensor at $\vx_i$, and the second term vanishes when $i=j$.

@@

%\listofchanges

>>>>

\usepackage{CJK}
\usepackage{microtype}
\usepackage{savesym}
  \savesymbol{lambdabar}
\usepackage{newpxtext, eulerpx}
  \restoresymbol{newpx}{lambdabar}
\usepackage[T1]{fontenc}
\usepackage{fontawesome}
\usepackage{amsmath}
%  \allowdisplaybreaks
\usepackage{bm}
\usepackage{nicefrac}
\usepackage{booktabs}
\usepackage[caption=false]{subfig}
\usepackage[figure, figure*]{hypcap}
\usepackage{tikz}
  \usetikzlibrary{positioning, fit, calc, arrows.meta}
\usepackage{framed}

@@

\begin{framed}
\vspace{1em}
\section*{Hamiltonian Dynamics}

Consider a Hamiltonian system, with $\vz = (\vx, \vp)$ where $\vx$ and
$\vp$ are the canonical coordinates and momenta.
Other variables split likewise: $\vDelta = (\vDelta_\vx, \vDelta_\vp)$,
$\vLambda = (\vLambda_\vx, \vLambda_\vp)$, and $\vlambda =
(\vlambda_\vx, \vlambda_\vp)$.
Hamilton's equation of motion is
%
\begin{equation}
\vf = \begin{pmatrix}
  \hphantom{-} \p\cH/\p \vp \\ - \p\cH/\p \vx
\end{pmatrix}.
\label{Hamilton}
\end{equation}
%
The dynamics is time-reversible: flipping the arrow of time $\d/\d t \to
- \d/\d t$ and the direction of momenta $\vp \to -\vp$ preserve the form
of \eqref{ode}.
It can be reversible after discretization if appropriate integrator is
used.

The variational equation \eqref{variational} and the adjoint equation
\eqref{adjoint} become
%
\begin{align}
\begin{pmatrix}
  \dot\vDelta_{\vx\vx} & \dot\vDelta_{\vx\vp} \\
  \dot\vDelta_{\vp\vx} & \dot\vDelta_{\vp\vp}
\end{pmatrix}
&= \begin{pmatrix}
  \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vx}
  & \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vp} \\
  - \dfrac{\p^2\cH}{\p\vx\p\vx}
  & - \dfrac{\p^2\cH}{\p\vx\p\vp} \\
\end{pmatrix}
\cdot
\begin{pmatrix}
  \vDelta_{\vx\vx} & \vDelta_{\vx\vp} \\
  \vDelta_{\vp\vx} & \vDelta_{\vp\vp}
\end{pmatrix}, \nonumber \\
%
\qquad
- \begin{pmatrix}
  \dot\vLambda_{\vx\vx} & \dot\vLambda_{\vx\vp} \\
  \dot\vLambda_{\vp\vx} & \dot\vLambda_{\vp\vp}
\end{pmatrix}
&= \begin{pmatrix}
  \hphantom{-} \dfrac{\p^2\cH}{\p\vx\p\vp}
  & - \dfrac{\p^2\cH}{\p\vx\p\vx} \\
  \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vp}
  & - \dfrac{\p^2\cH}{\p\vp\p\vx} \\
\end{pmatrix}
\cdot
\begin{pmatrix}
  \vLambda_{\vx\vx} & \vLambda_{\vx\vp} \\
  \vLambda_{\vp\vx} & \vLambda_{\vp\vp}
\end{pmatrix}.
\end{align}
%
Interestingly, their forms are identical once we swap rows of
$\vLambda$, and flip the directions of time and momenta.

As discussed before, $\vlambda$ follows adjoint equations similar to
that of $\vLambda$ above.
One can integrate it and \eqref{Hamilton} backward together to compute
$\d\cJ / \d\vtheta$.

\vspace{1em}
\section*{\texorpdfstring{$N$}{N}-body Dynamics}

We model self-gravitating collisionless dark matter with $N$ particles
each of mass $m$ in a periodic box of comoving volume $V$, whose
physical length expands proportional to a scale factor $a(t)$.
The canonical coordinates are the comoving positions $\vx = (\vx_1,
\cdots, \vx_N)$, and their canonical momenta are $\vp = (\vp_1, \cdots,
\vp_N)$ with $\vp_i = a^2 m \dot\vx_i$ for each particle.

The $N$-body Hamiltonian is
%
\begin{equation}
\cH = \sum_i \frac{\vp_i^2}{2 m a^2}
+ \sum_{i<j} \frac{m^2 \varphi(\vx_i - \vx_j)}{a},
\end{equation}
%
where $\varphi$ is the potential of a unit-mass particle at $a=1$ under
periodic boundary condition
%
\begin{equation}
\nabla^2 \varphi
= 4\pi G \bigl[ \deltaD(\vx - \vx') - V^{-1} \bigr].
\end{equation}
%
So the only nonzero second $\cH$ derivatives are
%
\begin{equation}
\frac{\p^2\cH}{\p\vp_i\p\vp_j} = \frac{\vI_3}{2 m a^2},
\qquad
\frac{\p^2\cH}{\p\vx_i\p\vx_j}
= \deltaK_{ij} \sum_{k \neq i} \frac{m^2}a
  \frac{\p\varphi(\vx_i-\vx_k)}{\p\vx_i\p\vx_i}
+ \frac{m^2}a \frac{\p\varphi(\vx_i-\vx_j)}{\p\vx_i\p\vx_j},
\end{equation}
%
where $\vI_3$ is the $3\times3$ identity matrix, the sum is the tidal
tensor at $\vx_i$, and the second term vanishes when $i=j$.
\end{framed}

@@
